Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-17 paper.qmd
Moved two nonlinear data code to script

```{r}
#| label: two-curvy-training
training_data_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_data.rds") |>
  mutate(ID = row_number())
```

```{r}
#| label: two-curvy-true-proj-data

true_model_df <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_true_model.rds")
wireframe_true_model <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_true_model_connections.rds")

data_two_curvy <- training_data_two_curvy |>
  select(-ID) |>
  mutate(type = "data")

true_model_two_curvy <- true_model_df |> 
  mutate(type = "true model")
```

<!--tSNE applied for Two curvy data-->
```{r}
#| label: tsne-two-curvy
tsne_two_curvy <- read_rds(file = "data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds") |>
  mutate(ID = row_number()) |>
  rename(c("emb1" = "tSNE1",
           "emb2" = "tSNE2"))
```

@@

num_bins_x_two_curvy <- 24

algo_obj_two_curvy <- fit_highd_model(
  highd_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  bin1 = num_bins_x_two_curvy, 
  q = 0.1, 
  benchmark_highdens = 1)

tsne_two_curvy_scaled <- algo_obj_two_curvy$nldr_obj$scaled_nldr
tr_from_to_df_two_curvy <- algo_obj_two_curvy$trimesh_data
df_bin_centroids_two_curvy <- algo_obj_two_curvy$model_2d
df_bin_two_curvy <- algo_obj_two_curvy$model_highd
hex_grid <- algo_obj_two_curvy$hb_obj$hex_poly
counts_df <- algo_obj_two_curvy$hb_obj$std_cts

hex_grid_with_counts <- left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hexID"))

hex_grid_nonempty <- hex_grid |>
  filter(hex_poly_id %in% df_bin_centroids_two_curvy$hexID)

bin_width <- algo_obj_two_curvy$hb_obj$a1

@@

# Second plot (identical to base except for xlim/ylim)
nldr_two_curvy2 <- nldr_two_curvy_original +
  interior_annotation("a", sc_ltr_pos, cex = 1.5)

# Third plot with aspect ratio and different annotation
nldr_two_curvy_original2 <- nldr_two_curvy_original +
  theme(aspect.ratio = 1) +
  interior_annotation("a1", sc_ltr_pos)

@@

## Compute error
error_df_two_curvy_abs <- augment(
  highd_data = training_data_two_curvy,
  model_2d = df_bin_centroids_two_curvy,
  model_highd = df_bin_two_curvy)

error_df_two_curvy_abs <- error_df_two_curvy_abs |>
  mutate(sqrt_row_wise_total_error = sqrt(row_wise_total_error))

error_df_two_curvy_abs <- error_df_two_curvy_abs |>
  bind_cols(tsne_two_curvy_scaled |>
              select(-ID))

@@

```{r}
#| label: prep-data-for-proj-tsne-best

df_bin_two_curvy <- df_bin_two_curvy |>
  select(-hexID) |>
  mutate(type = "model")

# Apply the scaling
df_model_data_two_curvy <- bind_rows(data_two_curvy, true_model_two_curvy, df_bin_two_curvy)

scaled_two_curvy <- scale_data_manual(df_model_data_two_curvy, "type") |>
  as_tibble()

scaled_two_curvy_data <- scaled_two_curvy |>
  filter(type == "data") |>
  select(-type)

scaled_two_curvy_data_true_model <- scaled_two_curvy |>
  filter(type == "true model") |>
  select(-type)

scaled_two_curvy_data_model <- scaled_two_curvy |>
  filter(type == "model") |>
  select(-type)

```{r}
#| label: langevitour-true-two-curvy-proj
#| eval: false
df_model_data_two_curvy_filtered <- bind_rows(true_model_two_curvy, data_two_curvy)

langevitour::langevitour(df_model_data_two_curvy_filtered[1:(length(df_model_data_two_curvy_filtered)-1)],
                         lineFrom = wireframe_true_model$from,
                         lineTo = wireframe_true_model$to,
                         group = factor(df_model_data_two_curvy_filtered$type,
                                        c("data", "true model")),
                         levelColors = c(clr_choice, "#000000"))
```

```{r}
#| label: langevitour-best-fit-two-curvy-proj-tsne
#| eval: false

df_model_data_two_curvy_filtered <- bind_rows(df_bin_two_curvy, data_two_curvy)

langevitour::langevitour(df_model_data_two_curvy_filtered[1:(length(df_model_data_two_curvy_filtered)-1)],
                         lineFrom = tr_from_to_df_two_curvy$from,
                         lineTo = tr_from_to_df_two_curvy$to,
                         group = factor(df_model_data_two_curvy_filtered$type,
                                        c("data", "model")),
                         levelColors = c(clr_choice, "#000000"))
```

@@

## First projection
model_prj1 <- cbind(
  c(0.37847,-0.06517,0.04231,0.01655),
  c(-0.06466,-0.27052,0.22131,-0.15235))

proj_obj1 <- get_projection(projection = model_prj1, 
               proj_scale = 1, 
               highd_data = scaled_two_curvy_data, 
               model_highd = scaled_two_curvy_data_model, 
               trimesh_data = tr_from_to_df_two_curvy, 
               axis_param = list(limits = 0.5, 
                                 axis_scaled = 2,
                                 axis_pos_x = -0.5, 
                                 axis_pos_y = -0.5, 
                                 threshold = 0.03))

## For true model

projection_scaled <- model_prj1 * 1

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% projection_scaled

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj1_true <- proj_obj1
proj_obj1_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true <- plot_proj(
  proj_obj = proj_obj1_true, 
  point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
  line_param = c(0.5, 0.5, "black"), # linewidth, alpha
  plot_limits = c(-0.6, 0.6), 
  axis_text_size = 4, 
  is_category = FALSE) +
  interior_annotation(label = "a", cex = 1.5) +
  theme(aspect.ratio=1, 
        legend.position = "none")

# Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj1 * 2,
    limits = 0.9,
    axis_pos_x = -0.4,
    axis_pos_y = -0.4,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.05)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj1[["axes"]] <- axes
proj_obj1[["circle"]] <- circle

@@

## Second projection
model_prj2 <- cbind(
  c(-0.38281,-0.04870,0.00619,-0.02451),
  c(-0.01303,0.00634,-0.37421,0.09648))

proj_obj2 <- get_projection(projection = model_prj2, 
                            proj_scale = 1, 
                            highd_data = scaled_two_curvy_data, 
                            model_highd = scaled_two_curvy_data_model, 
                            trimesh_data = tr_from_to_df_two_curvy, 
                            axis_param = list(limits = 0.35, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.35, 
                                              axis_pos_y = -0.35, 
                                              threshold = 0.01))

projected_true_model <- as.matrix(scaled_two_curvy_data_true_model) %*% model_prj2

projected_true_model_df <- projected_true_model |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(ID = dplyr::row_number())

true_model_df_proj <- dplyr::left_join(
  wireframe_true_model, 
  projected_true_model_df, 
  by = c("from" = "ID"))

names(true_model_df_proj)[3:NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_from")

true_model_df_proj <- dplyr::left_join(true_model_df_proj, projected_true_model_df, by = c("to" = "ID"))
names(true_model_df_proj)[(2 + NCOL(projected_true_model_df)):NCOL(true_model_df_proj)] <- paste0(names(projected_true_model_df)[-NCOL(projected_true_model_df)], "_to")

proj_obj2_true <- proj_obj2
proj_obj2_true[["model_df"]] <- true_model_df_proj

two_curvy_proj_true2 <- plot_proj(
    proj_obj = proj_obj2_true, 
    point_param = c(0.5, 0.5, clr_choice), # size, alpha, color
    line_param = c(0.5, 0.5, "black"), # linewidth, alpha
    plot_limits = c(-0.42, 0.42), 
    axis_text_size = 4, 
    is_category = FALSE) +
  interior_annotation(label = "b", cex = 1.5) +
  theme(aspect.ratio = 1, 
        legend.position = "none")

#Changed the axis parameters
axis_obj <- gen_axes(
    proj = model_prj2 * 2,
    limits = 0.7,
    axis_pos_x = -0.35,
    axis_pos_y = -0.35,
    axis_labels = names(scaled_two_curvy_data),
    threshold = 0.02)

axes <- axis_obj$axes
circle <- axis_obj$circle

proj_obj2[["axes"]] <- axes
proj_obj2[["circle"]] <- circle

@@

```{r}
#| label: code-illustration
# Code to draw illustration for notation
## hexagon binning to have regular hexagons
hb_obj_notation <- hex_binning(
  nldr_obj = algo_obj_two_curvy$nldr_obj, 
  bin1 = 7, 
  q = 0.1)

a1_temp <- hb_obj_notation$a1
a2_temp <- hb_obj_notation$a2
l_temp <- quad(a=3, b = 2 * a2_temp, c = -(a2_temp^2 + a1_temp^2))

## Data set with all centroids
all_centroids_df_temp <- hb_obj_notation$centroids
hex_grid_temp <- hb_obj_notation$hex_poly

hex_grid_temp40 <- hex_grid_temp |> 
  filter(hex_poly_id == 40)

start_pt <- all_centroids_df_temp |> 
  filter(hexID == 1)
d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = diff(algo_obj_two_curvy$nldr_obj$lim2)/diff(algo_obj_two_curvy$nldr_obj$lim1))

# To move the rectangle to ignore the overlap with the centroids
# rect_adj <- tibble(x1 = 0.03, x2 = 0.03)
rect_adj <- tibble(x1 = -0.03, x2 = 0.03)

a1 <- tibble(x = all_centroids_df_temp$c_x[4],
             xend = all_centroids_df_temp$c_x[5],
             y = all_centroids_df_temp$c_y[21],
             yend = all_centroids_df_temp$c_y[21],
             label = expression(a[1]))
a2 <- tibble(x = all_centroids_df_temp$c_x[25],
             xend = all_centroids_df_temp$c_x[25],
             y = all_centroids_df_temp$c_y[25],
             yend = all_centroids_df_temp$c_y[33],
             label = expression(a[2]))
l <- tibble(x = hex_grid_temp40$x[2],
            xend = hex_grid_temp40$x[3],
            y = hex_grid_temp40$y[2],
            yend = hex_grid_temp40$y[3],
            label = expression(l))

@@

num_bins_x_two_curvy <- 15

algo_obj_two_curvy2 <- fit_highd_model(
  highd_data = training_data_two_curvy, 
  nldr_data = tsne_two_curvy, 
  bin1 = num_bins_x_two_curvy, 
  q = 0.1, 
  benchmark_highdens = 5)

tr_from_to_df_two_curvy2 <- algo_obj_two_curvy2$trimesh_data
df_bin_centroids_two_curvy2 <- algo_obj_two_curvy2$model_2d
df_bin_two_curvy2 <- algo_obj_two_curvy2$model_highd

hex_grid_two_curvy2 <- algo_obj_two_curvy2$hb_obj$hex_poly
counts_df_two_curvy2 <- algo_obj_two_curvy2$hb_obj$std_cts

hex_grid_with_counts_two_curvy2 <- left_join(hex_grid_two_curvy2, counts_df_two_curvy2, by = c("hex_poly_id" = "hexID"))

@@

num_bins_x_two_curvy <- 48

algo_obj_two_curvy3 <- fit_highd_model(
    highd_data = training_data_two_curvy, 
    nldr_data = tsne_two_curvy, 
    bin1 = num_bins_x_two_curvy, 
    q = 0.1, 
    benchmark_highdens = 5)

tr_from_to_df_two_curvy3 <- algo_obj_two_curvy3$trimesh_data
df_bin_centroids_two_curvy3 <- algo_obj_two_curvy3$model_2d
df_bin_two_curvy3 <- algo_obj_two_curvy3$model_highd

hex_grid_two_curvy3 <- algo_obj_two_curvy3$hb_obj$hex_poly
counts_df_two_curvy3 <- algo_obj_two_curvy3$hb_obj$std_cts

hex_grid_with_counts_two_curvy3 <- left_join(hex_grid_two_curvy3, counts_df_two_curvy3, by = c("hex_poly_id" = "hexID"))

@@

#nldr_two_curvy2 + 
  trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 3)
```

@@

error_two_curvy_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")

## Find the minimum RMSE when have duplicate a1
error_two_curvy_umap <- error_two_curvy_umap |>
  group_by(a1) |>
  filter(bin1 == min(bin1)) |>
  ungroup() |>
  mutate(prop_dens = 1/(b*side_length^2))

base_line_prop_dens <- error_two_curvy_umap |>
  filter(a1 == min(a1)) |>
  pull(prop_dens)

error_two_curvy_umap <- error_two_curvy_umap |>
  mutate(prop_comp = prop_dens/base_line_prop_dens)

@@

error_two_curvy_umap <- error_two_curvy_umap |>
  mutate(prop_bins = b_non_empty/b)

@@

error_rm_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_rm_lwd_diff_bin.rds") |>
  mutate(bin1  = as.factor(bin1))

@@

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds")

@@

tsne_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_62.rds")

@@

umap_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

@@

phate_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_phate_knn_5.rds")

@@

trimap_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

@@

pacmap_two_curvy <- read_rds("data/two_non_linear_diff_shaped_close_clusters/two_non_linear_diff_shaped_close_clusters_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

@@

error_two_curvy_umap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_curvy_tsne <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_curvy_phate <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_curvy_trimap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_curvy_pacmap <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_curvy_tsne2 <- read_rds("data/two_non_linear_diff_shaped_close_clusters/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")

>>>>

@@

tsne_two_curvy_scaled <- read_rds("data/two_nonlinear/tsne_two_curvy_scaled.rds")
hex_grid_with_counts <- read_rds("data/two_nonlinear/hex_grid_with_counts.rds")
hex_grid_nonempty <- read_rds("data/two_nonlinear/hex_grid_nonempty.rds")
tr_from_to_df_two_curvy <- read_rds("data/two_nonlinear/tr_from_to_df_two_curvy.rds")
df_bin_centroids_two_curvy <- read_rds("data/two_nonlinear/df_bin_centroids_two_curvy.rds")

@@

@@

## Compute error
error_df_two_curvy_abs <- read_rds("data/two_nonlinear/error_df_two_curvy_abs.rds")

@@

@@

## First projection
proj_obj1 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj1.rds")

@@

proj_obj2 <- read_rds("data/two_nonlinear/two_nonlinear_proj_obj2.rds")

@@

```{r}
#| label: code-illustration
# Code to draw illustration for notation
## hexagon binning to have regular hexagons

hex_grid_temp <- read_rds("data/two_nonlinear/hex_grid_temp.rds")
all_centroids_df_temp <- read_rds("data/two_nonlinear/all_centroids_df_tempp.rds")
start_pt <- read_rds("data/two_nonlinear/start_pt.rds")
d_rect <- read_rds("data/two_nonlinear/d_rect.rds")
a1 <- read_rds("data/two_nonlinear/a2_data.rds")
a2 <- read_rds("data/two_nonlinear/a1_data.rds")
l <- read_rds("data/two_nonlinear/l_data.rds")
rect_adj <- read_rds("data/two_nonlinear/rect_adj.rds")

@@

hex_grid_with_counts_two_curvy2 <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy2.rds")
tr_from_to_df_two_curvy2 <- read_rds("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy2.rds")

@@

hex_grid_with_counts_two_curvy3 <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_with_counts_two_curvy3.rds")
tr_from_to_df_two_curvy3 <- read_rds("data/two_nonlinear/two_nonlinear_tr_from_to_df_two_curvy3.rds")

@@

trimesh_two_curvy_removed1_tsne_with_data +
  two_curvy_proj_tsne_first_model1 +
  two_curvy_proj_tsne_all_model1 +
  plot_layout(ncol = 3)
```

@@

error_two_curvy_umap <- read_rds("data/two_nonlinear/error_two_curvy_umap.rds")

@@

@@

error_rm_two_curvy <- read_rds("data/two_nonlinear/error_rm_lwd_diff_bin.rds") |>
  mutate(bin1  = as.factor(bin1))

@@

```{r}
#| label: read-two-curvy-clust-nldr
# Read a variety of different NLDR representations of two_non_linear_diff_shaped_close_clusters
# and plot them on same aspect ratio
tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_30.rds")

@@

tsne_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_tsne_perplexity_62.rds")

@@

umap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_umap_n-neigbors_15_min-dist_0.1.rds")

@@

phate_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_phate_knn_5.rds")

@@

trimap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_trimap_n-inliers_12_n-outliers_4_n-random_3.rds")

@@

pacmap_two_curvy <- read_rds("data/two_nonlinear/two_non_linear_diff_shaped_close_clusters_pacmap_n-neighbors_10_init_random_MN-ratio_0.5_FP-ratio_2.rds")

@@

error_two_curvy_umap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_umap.rds")
error_two_curvy_tsne <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne.rds")
error_two_curvy_phate <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_phate.rds")
error_two_curvy_trimap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_trimap.rds")
error_two_curvy_pacmap <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_pacmap.rds")
error_two_curvy_tsne2 <- read_rds("data/two_nonlinear/error_two_non_linear_diff_shaped_close_clusters_tsne2.rds")
